---
title: "Homework Assignment 1"
author: "Aditya Mittal"
date: 'Assigned: Oct 24, 2020, Due Sun Nov 01, 2020 11:59PM'
output:
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
  pdf_document: default
  
---

## Data frame basics

```{r}
nycflights <- read.csv("nycflights.csv")
#Removing garbage column X
nycflights$X <- NULL
#Changing the format of variables
nycflights$origin <- as.factor(nycflights$origin)
nycflights$tailnum <- as.factor(nycflights$tailnum)
nycflights$carrier <- as.factor(nycflights$carrier)
nycflights$month <- as.factor(nycflights$month)
nycflights$dest <- as.factor(nycflights$dest)
nycflights$flight <- as.factor(nycflights$flight)
nycflights$time_hour <- as.Date(nycflights$time_hour)
head(nycflights)

```

### Data Exploration

```{r}

#Number of Airlines/carriers
nlevels(nycflights$carrier)
#List of all those airlines
levels(nycflights$carrier)
#Number of flights by the airline with code OO
nrow(nycflights[nycflights$carrier == 'OO',])
#shortest flight out of any NYC airport
min(nycflights$air_time, na.rm = TRUE)
#flights by United Airlines (code: UA) on Jan 12th 2013
nrow(nycflights[nycflights$carrier == 'UA' & nycflights$time_hour == '2013-01-13',])
```



### Arrival Delay

```{r}
#average arrival delay for all airports and all airlines combined in Jan 2013
round(mean(nycflights[nycflights$month == '1',]$arr_delay, na.rm = TRUE),2)

#median arrival delay for all airports and all airlines combined in Jan 2013
round(median(nycflights[nycflights$month == '1',]$arr_delay, na.rm = TRUE),2)
```
Since mean arrival delay is greater than median arrival delay, arrival delay variable is positively skewed.


### Airline Performance

```{r}
tapply(nycflights$arr_delay, nycflights$carrier, mean, na.rm = TRUE)
Arrival_Delay_Byairline <- tapply(nycflights$arr_delay, nycflights$carrier, mean, na.rm = TRUE)
barplot(Arrival_Delay_Byairline)
```

The airline with the highest arrival delay is: `r names(Arrival_Delay_Byairline[Arrival_Delay_Byairline == max(Arrival_Delay_Byairline)])` and the amount of delay is: `r max(Arrival_Delay_Byairline)`
The airline with the smallest arrival delay is: `r names(Arrival_Delay_Byairline[Arrival_Delay_Byairline == min(Arrival_Delay_Byairline)])` and the amount of delay is: `r min(Arrival_Delay_Byairline)`
The airlines with negative arrival delays are: `r names(Arrival_Delay_Byairline[Arrival_Delay_Byairline <0])`


### Air Gain

```{r}
nycflights$airgain <- nycflights$dep_delay - nycflights$arr_delay
summary(nycflights$airgain)
tapply(nycflights$airgain, nycflights$origin, mean, na.rm = TRUE)
tapply(nycflights$airgain, nycflights$carrier, mean, na.rm = TRUE)
tapply(nycflights$airgain, nycflights$month, mean, na.rm = TRUE)
hist(nycflights$airgain)
barplot(tapply(nycflights$airgain, nycflights$origin, mean, na.rm = TRUE))
barplot(tapply(nycflights$airgain, nycflights$carrier, mean, na.rm = TRUE))
barplot(tapply(nycflights$airgain, nycflights$month, mean, na.rm = TRUE))
library("ggplot2")
ggplot(data=nycflights, aes(x=origin, y=airgain)) + geom_point()
ggplot(data=nycflights, aes(x=carrier, y=airgain)) + geom_point()
ggplot(data=nycflights, aes(x=month, y=airgain)) + geom_point()

#Do airlines actually gain any time when in air on average
mean(nycflights$airgain, na.rm = TRUE)
#Average airgain by different airlines
tapply(nycflights$airgain, nycflights$carrier, mean, na.rm = TRUE)

```


### Merging Data Frames

```{r}
planes <- read.csv("planes.csv")
#Removing garbage column X
planes$X <- NULL
#Changing the format of variables
planes$tailnum <- as.factor(planes$tailnum)
planes$type <- as.factor(planes$type)
planes$manufacturer <- as.factor(planes$manufacturer)
planes$engines <- as.factor(planes$engines)
planes$engine <- as.factor(planes$engine)
planes$model <- as.factor(planes$model)
dim(nycflights)
dim(planes)
library(dplyr)
nycflightsmerged <- left_join(nycflights, planes, by = c("tailnum" = "tailnum"), suffix = c(".x", ".y"), .merge = TRUE)
head(nycflightsmerged)
dim(nycflightsmerged)

#Q1 - How does average air gain differ by engine
barplot(tapply(nycflightsmerged$airgain,nycflightsmerged$engine, mean, na.rm = TRUE))

#Q2 - How does average air gain differ by type
barplot(tapply(nycflightsmerged$airgain, nycflightsmerged$type, mean, na.rm = TRUE))

# Q3 - What is oldest aircraft across carriers by year
tapply(nycflightsmerged$year.y,nycflightsmerged$carrier, min, na.rm = TRUE)

#Q4 - Which carrier flies the biggest plane in terms of number of seats
maxseats <- tapply(nycflightsmerged$seats,nycflightsmerged$carrier, max, na.rm = TRUE)
names(maxseats[maxseats == max(maxseats)])

#Q5 - what is the number of flights departed from JFK that was manufactured by BOEING
nrow(nycflightsmerged[nycflightsmerged$manufacturer == 'BOEING' & nycflightsmerged$origin=='JFK',])
```
